---
title: "Biogeographical analysis"
bibliography: refs.yaml
csl: cladistics.csl
format:
  docx:
    toc: false
    number-sections: false
    highlight-style: github
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)

source(here::here("R/packages.R"))
source(here::here("R/functions.R"))

number <- function(...) {
  scales::number(big.mark = ",", ...)
}

# Load objects from workflow
tar_load(
  c(
    anc_states_probs,
    bgb_stats_pretty,
    bsm_events,
    antro_phy),
  store = here::here("_targets")
  )
```

```{r}
#| label: antrophyopsis
#| include: false

# node 34 is Antrophyum
n_disp_rep_node_34 <-
  bsm_events %>%
  filter(node == 34, event_type == "d") %>%
  count(rep) %>%
  pull(rep) %>%
  n_distinct()
```

## Methods

### Historical biogeographic analysis

We used the chronogram resulting from BEAST as the input file for historical biogeographic analysis.
We determined geographic distributions for species based upon our own collections that we personally identified in order to avoid problems caused by potential misidentifications.
We recognized six areas including: (A) Africa, (B) Asia [Asia-temperate + Indian subcontinent + Indo-China], (C) Malesia, (D) Papuasia, (E) Pacific, and (F) Australasia, which follows the geographical scheme of Brummitt (2001).
We chose these areas with the aim to best represent the current global distribution of the genus and to detect past major biogeographical shifts within the genus.
We used the R package ‘BioGeoBEARS’ (Matzke, 2013) to perform a likelihood ancestral range estimation using three models: (1) DEC (Dispersal Extinction Cladogenesis, Ree and Smith, 2008), (2) DIVALIKE (a likelihood-based implementation of dispersal vicariance analysis, Ronquist, 1997); and (3) BAYAREALIKE (a likelihood implementation of BayArea, Landis et al., 2013).
We also tested models with and without founder-event speciation, which is incorporated with the “j” parameter in the package.
We further evaluated the models under a constrained analysis which considered the dispersal rate differences among the six areas.
Specifically, we set a dispersal probability matrix based on the connectivity among the bioregions (1 = high dispersal rate within an area, 0.5 = medium dispersal rate between connected areas, and 0.05 = low dispersal rate between disconnected areas, Table 2).
In total, we tested 12 models (Table 3) and selected the best-fit model based on the Akaike Information Criterion (AIC, Burnham and Anderson, 1998) since not all of the models are nested.

We used biogeographical stochastic mapping (BSM) to estimate the frequency of dispersal and vicariance events based on the best-fit model [@Dupin2017].
We simulated biogeographic events in *Antrophyum* using 1,000 replicates and summarized the mean and standard deviation (SD) of the frequency of events across replicates.
When analyzing temporal trends in the frequency of biogeographic events, we standardized by total branch length in bins of 2 my each, since the number of events scales with branch length (or nodes, in the case of cladogenic events) and the total number of branches/nodes increases towards the present [@Silvestro2018].

All code for biogeographical analyses are available at https://github.com/joelnitta/antrophyum_ow.

## Results

### Molecular dating and historical biogeography

```{r}
#| label: events-table
#| include: false

# format names for pretty table
pretty_event_names <- tribble(
  ~event_type, ~type, ~mode, ~short_name,
  "d", "Range expansion (d)", "Dispersal", "d",
  "founder (j)", "Founder (j)", "Dispersal", "j",
  "subset (s)", "Subset (s)", "Within-area speciation", "s",
  "sympatry (y)", "Sympatry (y)", "Within-area speciation", "y",
  "vicariance (v)", "Vicariance (v)", "Vicariance", "v"
) %>%
  mutate(
    mode = factor(
      mode, levels = c("Within-area speciation", "Dispersal", "Vicariance"))
  )

# pretty table for printing
bsm_mean_event_table <-
bsm_events %>%
  count_mean_bsm(event_type) %>%
  left_join(pretty_event_names, by = "event_type") %>%
  arrange(mode, type) %>%
  # select(mode, type, mean, sd) %>%
  select(Mode = mode, Event = type, Mean = mean, SD = sd)

# list for referencing in-line
bsm_counts <-
bsm_events %>%
  count_mean_bsm(event_type) %>%
  left_join(pretty_event_names, by = "event_type") %>%
  mutate(
    mean = round(mean, 2),
    sd = round(sd, 2)) %>%
  split(.$short_name)
```

The chronogram inferred by BEAST with the ancestral range estimated by BioGeoBEARS is summarized in Fig. 8.
The tree topology derived from BEAST was identical to that from RAxML.
The divergence of *Antrophyum* and its sister genus Antrophyopsis was dated back to the middle Oligocene around 30 Ma (95% highest posterior density, HPD = 24.0–36.1 Ma).
Within the genus, the first and second divergences dated back to around 20 Ma (95% HPD = 15.3–25.6 Ma) and 16 Ma (95% HPD = 11.8–20.3 Ma).

```{r}
#| label: find-best-mod
#| include: false

# Extract best-scoring model name
best_mod_basic <- bgb_stats_pretty %>%
  slice_min(aic) %>%
  pull(mod_name_pretty)

# Extract constraint type
best_mod_constraint <- bgb_stats_pretty %>%
  slice_min(aic) %>%
  pull(constraint)

# Combine to get name of best-scoring model
best_mod <- paste(best_mod_constraint, best_mod_basic)
```


Among the 12 biogeographic models tested, we determined the `r best_mod` model as the best-fit model because it yielded the lowest AIC value (Table 3).
Table 8 shows the most likely ancestral range of each clade, together with divergence-time estimates.
The best-fit model suggested the ancestor of *Antrophyum* was widely distributed in Malesia, Papuasia, and Pacific Islands.
Nonetheless, there is high uncertainty in ancestral ranges (Fig. S17).
Papuasian and Malesian origins were inferred for Clades B and C, respectively.
These clades began to expand into other areas about 10 Ma.

Range expansions were the most frequently observed biogeographic event over all BSM replicates (mean `r bsm_counts$d$mean` ± `r bsm_counts$d$sd` events per replicate; Table X).
The most frequently observed dispersal patterns were between Malesia and Papuasia, from Papuasia to the Pacific, and between Asia and Malesia (Fig. S18A).
Vicariance events were most frequent early in the history of *Antrophyum*, and decreased with time towards the present (Fig. S19A).
Dispseral events increased in the most recent 10 my (Fig. S19).

## Discussion

Our study presents the first comprehensive phylogenetic analysis of *Antrophyum*. 
Among the 34 recognized species, only three of them are missing in the current analysis and it is unlikely they will affect the main conclusions (see later discussion). 
The time-calibrated phylogeny reveals that the common ancestor of *Antrophyum* and its sister group *Antrophyopsis* diverged in the middle Oligocene around 30 Ma (95% HPD = 23.9–36.1 Ma).
The current allopatric distribution of *Antrophyum* (mainly tropical Asia and Pacific Islands) and Antrophyopsis (African endemic) is likely the result of transoceanic dispersal since the two landmasses were already separated at that time; at least one dispersal event was observed in `r n_disp_rep_node_34` of the 1,000 BSM replicates along the branch leading to *Antrophyum*.
Dispersal seems to have played an important role in the biogeographic history of *Antrophyum*, with an average of `r bsm_counts$j$mean + bsm_counts$d$mean` dispersal events (range expansions or founder events) per BSM replicate, but only `r bsm_counts$v$mean` vicariance events (Table X).

Although the estimation of ancestral ranges under the best-fit model (constrained DEC + j) suggests that the ancestral range of crown group of *Antrophyum* to be on the landmasses of today’s Malesia (C), Papuasia (D), and Pacific (E), the low probability of this result (15%, Table 8) leaves the ancestral distribution of the genus uncertain.
Unexpectedly, the New Caledonian endemic A. novae-caledoniae is recovered as sister to all other living species.
Their divergence was dated to the early Miocene around 20 Ma (95% HPD = 15.3–25.6 Ma).
A similar pattern has been also observed in several other plant groups (Álvarez-Molina & Cameron, 2009; Vasconcelos et al., 2017) and the most well-known example is the flowering plant *Amborella*, which is endemic to New Caledonia and sister to all other angiosperms (Simmons 2017).
So-called relict lineages seem to be over-represented in New Caledonia, and this has been attributed to the persistence of forest throughout the Neogene that acted as refugia while extinctions happened elsewhere (Pouteau et al. 2015, Condamine et al. 2017, Pillon et al. 2021).

```{r}
#| label: clade_b_c
#| include: false

# Calculate mean number of dispersals from Papuasia (D) to Malesia (C)

b2_d2c_events <-
bsm_events %>%
  # dispersal includes range expansion and founder
  mutate(
    event_type = fct_collapse(
      event_type,
      dispersal = c("d", "founder (j)"))
  ) %>%
  # Papuasia (D) to Malesia (C)
  filter(str_detect(from, "D")) %>%
  filter(str_detect(to, "C")) %>%
  filter_by_clade_and_count(
    antro_phy,
    # Clade B-2
    c("V363_A_plantagineum", "V549_A_immersum"),
    .) %>%
  split(.$event_type)

b3_d2c_events <-
bsm_events %>%
  # dispersal includes range expansion and founder
  mutate(
    event_type = fct_collapse(
      event_type,
      dispersal = c("d", "founder (j)"))
  ) %>%
  # Papuasia (D) to Malesia (C)
  filter(str_detect(from, "D")) %>%
  filter(str_detect(to, "C")) %>%
  filter_by_clade_and_count(
    antro_phy,
    # Clade B-3
    c("V386_A_solomonense", "V464_A_strictum"),
    .) %>%
  split(.$event_type)
```

Clades B and C include the majority of living species, and there is a clear difference in their geography: the origin of clade B was estimated to be Papuasia whereas that of clade C was Malesia (Table 8).
Clade B subsequently colonized eastward to the Pacific Islands and westward to Malesia while the clade C colonized northward to Asia (Fig. S18B, C).
The extent to which the contrasting biogeographic histories of clades B and C reflect ecological differences remains to be investigated.
Long distance dispersal, in particular range expansions, increased in frequency during the past 10 Ma (Fig. S19B).
Although the overall trend is to disperse eastward (from Papuasia or Malesia to the Pacific Islands) or northward (from Malesia to Asia), the analysis also suggests several westward dispersal events in clade B.
For example, the B-2 and B-3 clades dispersed from Papuasia to Malesia (mean `r b2_d2c_events$dispersal$mean` ± `r b2_d2c_events$dispersal$sd` and `r b3_d2c_events$dispersal$mean` ± `r b3_d2c_events$dispersal$sd` such dispersal events per BSM replicate, respectively).

```{r}
#| label: long-dist-io
#| include: false
immersum_c2a_events <-
bsm_events %>%
  # dispersal includes range expansion and founder
  mutate(
    event_type = fct_collapse(
      event_type,
      dispersal = c("d", "founder (j)"))
  ) %>%
  # Malesia (C) to Africa (A)
  filter(str_detect(from, "C")) %>%
  filter(str_detect(to, "A")) %>%
  filter_by_clade_and_count(antro_phy, "V549_A_immersum", .) %>%
  split(.$event_type)

malgassicum_c2a_events <-
bsm_events %>%
  mutate(
    event_type = fct_collapse(
      event_type,
      dispersal = c("d", "founder (j)"))
  ) %>%
  # Malesia (C) to Africa (A)
  filter(str_detect(from, "C")) %>%
  filter(str_detect(to, "A")) %>%
  filter_by_clade_and_count(
    antro_phy,
    c("V360_A_sessilifolium", "V586_A_malgassicum"),
    .) %>%
  split(.$event_type)

```

Long-distance dispersal that crossed the Indian Ocean (from Malesia to Africa) likely happened at least twice, once in *A. immersum* and the other in the clade containing *A. sessifolium*, *A. malgassicum* and closely related species (mean `r immersum_c2a_events$dispersal$mean` ± `r immersum_c2a_events$dispersal$sd` and `r malgassicum_c2a_events$dispersal$mean` ± `r malgassicum_c2a_events$dispersal$sd` such dispersal events per BSM replicate, respectively).

## Tables

Table 3. Results of the BioGeoBEARS analysis. LnL, AIC, and model parameters rounded to two decimal points.

```{r}
#| label: bsm-table
#| echo: false

bgb_stats_pretty %>%
  mutate(
    across(c(lnl, d, e, j, aic), ~round(., 2))
  ) %>%
  select(
    Model = mod_name_pretty,
    `Geographical constraint` = constraint,
    `LnL` = lnl,
    `No. of parameters` = numparams,
    d, e, j,
    AIC = aic
  ) %>%
  knitr::kable()
```

Table 8. Bayesian posterior divergence-time estimates (Median and 95% highest posterior density) for the main clades of Antrophyum, along with the most probable ancestral ranges.

```{r}
#| label: prob-table
#| echo: false

# Ages copied from original MS
ages_table <- tribble(
  ~stem_age, ~crown_age, ~clade,
  "29.9 (23.9–36.1)", "20.2 (15.3–25.6)", "*Antrophyum*",
  "20.2 (15.3–25.6)", "16.1 (11.8–20.3)", "B + C",
  "16.1 (11.8–20.3)", "12.8 (9.1–16.7)", "B",
  "16.1 (11.8–20.3)", "13.5 (10.0–18.0)", "C"
)

node_table <- tribble(
  ~node, ~clade,
  34, "*Antrophyum*",
  35, "B + C",
  53, "B",
  36, "C"
) %>%
  left_join(ages_table, by = "clade") %>%
  mutate(clade = factor(clade, levels = c("*Antrophyum*", "B + C", "B", "C")))

anc_states_probs %>%
  pivot_longer(names_to = "range", values_to = "prob", -node) %>%
  inner_join(node_table, by = "node") %>%
  group_by(node) %>%
  slice_max(prob) %>%
  ungroup() %>%
  arrange(clade) %>%
  transmute(
    Clade = clade,
    `Stem group age in Ma (95% HPD)` = stem_age,
    `Crown group age in Ma (95% HPD)` = crown_age,
    `Most probable ancestral ranges` = paste0(
      range, " (", round(prob * 100, 0), "%)")
  ) %>%
  knitr::kable()
```

Table X: Frequency of biogeographic events observed over 1,000 biogeographical stochastic mapping (BSM) replicates using best-scoring model (`r best_mod`). Letters in parentheses indicate name of parameter in BioGeoBEARS model. No extinction events were observed. 

```{r}
#| label: bsm-event-table
#| echo: false

bsm_mean_event_table %>%
  knitr::kable()
```

## Supplemental Figures

```{r}
#| label: dispersal-bars
#| include: false

# Define function to make plot
plot_event_count <- function(event_counts) {
  event_counts %>%
    mutate(event_txt = fct_reorder(event_txt, mean)) %>%
    ggplot(aes(x = mean, y = event_txt)) +
    geom_errorbarh(aes(xmin = mean - 0.01, xmax = mean + sd, height = 0.2)) +
    geom_col() +
    scale_x_continuous(name = "Frequency") +
    theme(axis.title.y = element_blank())
}

bsm_disperals <-
  bsm_events %>%
  filter(event_type %in% c("founder (j)", "d")) %>%
  count_mean_bsm(event_txt) %>%
  filter(mean > 0.5)

# To double check node numbers:
# plot(ladderize(antro_phy, right = FALSE))
# nodelabels()
# tiplabels()

# Don't include crown node because disperals are on branch leading
# to node
b_nodes <- phangorn::Descendants(antro_phy, 53, "all")
c_nodes <- phangorn::Descendants(antro_phy, 36, "all")

clade_b_dispersals <-
bsm_events %>%
  filter(node %in% b_nodes) %>%
  filter(event_type %in% c("founder (j)", "d")) %>%
  count_mean_bsm(event_txt) %>%
  filter(mean > 0.5)

clade_c_dispersals <-
bsm_events %>%
  filter(node %in% c_nodes) %>%
  filter(event_type %in% c("founder (j)", "d")) %>%
  count_mean_bsm(event_txt) %>%
  filter(mean > 0.5)

fig_s18a <- plot_event_count(bsm_disperals) +
  ggtitle("Full tree")

fig_s18b <- plot_event_count(clade_b_dispersals) +
  ggtitle("Clade B")

fig_s18c <- plot_event_count(clade_c_dispersals) +
  ggtitle("Clade C")

fig_s18 <- fig_s18a + fig_s18b + fig_s18c + 
  plot_spacer() +
  plot_annotation(tag_levels = "A")

ggsave(
  plot = fig_s18,
  file = here::here("reports/FigureS18_bsm_dispersals.pdf"),
  height = 9, width = 6.5)

```

Figure S18: Frequency (mean and SD) of dispersal events (range expansion and founder events) across 1,000 biogeographical stochastic mapping (BSM) replicates. Only events with at least mean of 0.5 occurrences per replicate shown. Region codes as in Figure 8. Arrow indicates direction of movement. Comma indicates a founder event; other events are range expansions. (A) full tree, (B) clade B, (C) clade C.

```{r}
#| label: events-over-time
#| include: false

# Plot events over time, normalized by branchlength in 2 my bins

## helper function for plotting

# plot of normalized dispersal events through time
plot_normalized_events <- function(normalized_data) {
  ggplot(normalized_data, aes(x = time, ymin = mean - sd, ymax = mean + sd)) +
    pammtools::geom_stepribbon(fill = "grey70") +
    geom_step(aes(y = mean), direction = "hv") +
    xlab("Time (mya)") +
    ylab("Frequency") +
    scale_x_reverse()
}

## get normalized counts of events ----

# bin events and branchlenths
# - set bin width
bin_width <- 2
# - bin branch lengths
blrn_binned <- bin_brln(antro_phy, bin_width)
# - all dispersal events (range expansion + founder events)
disp_binned <- bin_events(bsm_events, c("d", "founder (j)"), bin_width)
# - only founder events
founder_binned <- bin_events(bsm_events, "founder (j)", bin_width)
# - only range expansions
range_exp_binned <- bin_events(bsm_events, "d", bin_width)
# - vicariance events
vic_binned <- bin_events(bsm_events, "vicariance (v)", bin_width)

disp_normalized <- normalize_events(disp_binned, blrn_binned)
founder_normalized <- normalize_events(founder_binned, blrn_binned)
range_exp_normalized <- normalize_events(range_exp_binned, blrn_binned)
vic_normalized <- normalize_events(vic_binned, blrn_binned)

fig_s19a <- plot_normalized_events(vic_normalized) +
  ggtitle("Vicariance events")

fig_s19b <- plot_normalized_events(range_exp_normalized) +
  ggtitle("Range expansion events")

fig_s19c <- plot_normalized_events(founder_normalized) +
  ggtitle("Founder events")

fig_s19d <- plot_normalized_events(disp_normalized) +
  ggtitle("All dispersal events")

fig_s19 <- fig_s19a + fig_s19b + fig_s19c + fig_s19d +
  plot_annotation(tag_levels = "A")

ggsave(
  plot = fig_s19,
  file = here::here("reports/FigureS19_bsm_events_over_time.pdf"),
  height = 7, width = 6.5)
```

Figure S19: Frequency of biogeographic events over time. Y-axis shows mean (stepped lines) and SD (grey box) from 1,000 biogeographical stochastic mapping (BSM) replicates normalized by total branch length in 2 my bins. No data plotted for bins with zero events observed over all replicates. (A) vicariance events, (B) range expansion events, (C) founder events, (D) all dispersal events (range expansion and founder events).

## References
